---
title: "March Madness Upsets"
author: "Cooper Olney"
date: "2026-04-13"
format: 
  html:
    embed-resources: true
---

# Abstract



# Introduction
The NCAA Men's Basketball Tournament, commonly known as March Madness, is one of the most widely followed and unpredictable sporting events in the United States. It features 68 teams in a single elimination format, it includes many of the most well known teams in the country and some lesser known teams. This structures creates unexpected outcomes, making March Madness especially exciting for fans filling out brackets, however no one has ever achieved a perfect bracket in the history of the tournament.

One of the most defining features of March Madness is the occurrence of upsets, in which a lower seeded teams defeats a higher seeded team, for this data the seeds need to differ by at least three spots. Higher sees are generally expected to perform better, but the tournament consistently produces surprising results. The lower seeded teams that make deep runs in the tournament are often referred to as "Cinderella teams". These teams are a major reason why March Madness remains so exciting and difficult to predict.

The data used in this analysis comes two datasets (dev & upset) sourced from Kaggle. Upset tracks the data on the number of upsets per tournament, the round the upset occurred and the seeds of the teams in those games. dev is a collection of different on the court metrics and other off the court statistics. Together, these variables off insight... (clean this up)

**Main Research Question**
What team characteristics are associated with unexpected success in March Madness?

To address this question, this analysis explores several related areas. First, it examines how common upsets are in order to provide context for how frequently Cinderella teams actually are. It then investigates which on-court performance metrics are most associated with unexpected success. In addition, the analysis considers some off-court factors like team experience and physical attributes. Finally, famous Cinderella teams like St. Peters, Loyola Chicago, and Davidson are analyzed to determine whether they follow similar patterns.

Some additional context about the structure of the tournament is important for interpreting these results. Since March Madness is single elimination it increasing the role of variability and short term performance. Additionally, games are played at neutral sites, which reduce advantages what occur in the regular season. These factors contribute to the unpredictability of outcomes and make it more challenging to identify consistent predictors of success.

However, this analysis will determine if there are patterns in the unexpected success in the tournament. Overall, these factors motivate a deeper exploration into the statistical and contextual forces of upsets and Cinderella runs in March Madness. (clean this up)


# Visualizations
```{r}
library(tidyverse)
```

```{r}
# Cleaning datasets

# for plot 1
upset_summary <- upset_seed_info |> group_by(YEAR) |>
  summarise(num_upsets = n(),
            avg_seed_diff = round(mean(`SEED DIFF`, na.rm=TRUE), 3))

# for plot 2
on_court_performance <- dev |> filter(`Post-Season Tournament Sorting Index` == 1) |>
  select(
    Season, `Mapped ESPN Team Name`, Seed, 
    `Net Rating`, `Adjusted Tempo`,
    `Adjusted Offensive Efficiency`, `Avg Possession Length (Offense)`, eFGPct, TOPct, ORPct, FTRate, Off3PtFG, FTPct, Bench, 
    `Adjusted Defensive Efficiency`, `Avg Possession Length (Defense)`, DefFT, Off3PtFG, 
         ) |>
  mutate(Seed = as.numeric(Seed)) |>
  filter(Seed >= 10) |>
  mutate(cinderella = 0)
  # create a cinderella vs non-cinderella team so 10 >= and advanced to sweet 16
# need to find all 10+ seeds to make sweet 16 since 2002

write_csv(on_court_performance, "on_court_performance.csv")

library(readxl)
data <- read_excel("data/new_mm_datasets/on_court_performance.xlsx")
# bench and possession times have N/A values so maybe dont use them, so remove them for now
# current on court performance dataset
data <- data |> select(-c(`Avg Possession Length (Offense)`, `Avg Possession Length (Defense)`, Bench)) |>
  relocate(cinderella, .after = Seed) |>
  mutate(cinderella = if_else(cinderella == 1, "Yes", "No"))
  
# create cinderella var first
boxplot_data <- data %>%
  select(
    cinderella,
    `Net Rating`,
    `Adjusted Offensive Efficiency`,
    `Adjusted Defensive Efficiency`,
    ORPct,
    TOPct,
    `Adjusted Tempo`
  ) %>%
  pivot_longer(-cinderella, names_to = "metric", values_to = "value")

# for plot 3
# merge it with data to get the other columns back
# columns to keep
# other data to look for: age of players, team health, travel distance
# this data is NA before 2007
off_court_factors <- dev |> filter(`Post-Season Tournament Sorting Index` == 1) |>
  select(
    Season, `Mapped ESPN Team Name`, Seed, `Short Conference Name`,
    AvgHeight, CenterHeight, PFHeight, SFHeight, SGHeight, PGHeight, EffectiveHeight, Experience, `Active Coaching Length` 
         ) |>
  mutate(Seed = as.numeric(Seed)) |>
  filter(Seed >= 10) |>
  filter(Season >= 2007)

# helper datasets
cinderella_var_data <- data %>%
  select(Season, `Mapped ESPN Team Name`, cinderella)

coferences <- off_court_factors |> group_by(`Short Conference Name`) |> summarise(n = n())

power_confs <- c("ACC", "B10", "B12", "SEC", "P12", "BE")

# continue cleaning plot 3 data
off_court_factors <- off_court_factors %>%
  left_join(cinderella_var_data, by = c("Season", "Mapped ESPN Team Name")) |>
  relocate(cinderella, .after = Seed) |>
  arrange(desc(Season)) |>
  mutate(
    `Active Coaching Length` = na_if(`Active Coaching Length`, "Unknown")
    ) |>
  mutate(
    `Active Coaching Length` = gsub(" years", "", `Active Coaching Length`),
    `Active Coaching Length` = gsub(" year", "", `Active Coaching Length`),
    `Active Coaching Length` = as.numeric(`Active Coaching Length`)
  ) |>
  mutate(
    `Active Coaching Length` = if_else(`Mapped ESPN Team Name` == "Yale",
                                       27,
                                       `Active Coaching Length`),
    `Active Coaching Length` = if_else(`Mapped ESPN Team Name` == "Gonzaga",
                                       27,
                                       `Active Coaching Length`)
  ) |>
  mutate(cinderella = if_else(cinderella == 1, "Yes", "No")) |>
  mutate(
    conference_type = if_else(`Short Conference Name` %in% power_confs,
                              "Power",
                              "Mid-Major")
  ) |>
  relocate(conference_type, .after = `Short Conference Name`)

plot3_data <- off_court_factors |>
  select(
    cinderella,
    conference_type,
    Experience,
    EffectiveHeight,
  ) %>%
  pivot_longer(cols = -c(cinderella, conference_type), 
               names_to = "metric", 
               values_to = "value")


# Plot 4
# merge on court and off court
combined_data <- inner_join(off_court_factors, data,
           by = c("Season", "Mapped ESPN Team Name", "Seed", "cinderella"))

scaled_data <- combined_data %>%
  mutate(
    NetRating_z = scale(`Net Rating`),
    OffEff_z = scale(`Adjusted Offensive Efficiency`),
    DefEff_z = scale(`Adjusted Defensive Efficiency`),
    Experience_z = scale(Experience),
    EffHeight_z = scale(EffectiveHeight)
  )

avg_cinderella <- scaled_data %>%
  filter(cinderella == "Yes") %>%
  summarise(
    team = "Avg Cinderella",
    NetRating_z = mean(NetRating_z, na.rm = TRUE),
    OffEff_z = mean(OffEff_z, na.rm = TRUE),
    DefEff_z = mean(DefEff_z, na.rm = TRUE),
    Experience_z = mean(Experience_z, na.rm = TRUE),
    EffHeight_z = mean(EffHeight_z, na.rm = TRUE)
  )

case_teams <- scaled_data %>%
  filter(
    (`Mapped ESPN Team Name` == "Davidson" & Season == 2008) |
    (`Mapped ESPN Team Name` == "Saint Peter's" & Season == 2022) |
    (`Mapped ESPN Team Name` == "Loyola Chicago" & Season == 2018)
  ) %>%
  select(
    team = `Mapped ESPN Team Name`,
    NetRating_z, OffEff_z, DefEff_z, Experience_z, EffHeight_z
  )

plot4_data <- bind_rows(case_teams, avg_cinderella) |>
  pivot_longer(-team, names_to = "metric", values_to = "value")
  
```

## Plot 1: March Madness Upsets Over Time

To understand how common upsets are in March Madness, we can look at the following lollipop plot, which shows the number of upsets in each tournament.
```{r}
ggplot(upset_summary, aes(x = YEAR, y = num_upsets)) +
  geom_segment(aes(x = YEAR, xend = YEAR, y = 0, yend = num_upsets,
                   color = avg_seed_diff)) +
  geom_point(aes(color = avg_seed_diff), size = 2) +
  scale_color_viridis_c(name ="Avg Seed Diff") +
  coord_flip() +
  labs(
    title = "March Madness Upsets Over Time",
    x = "Year",
    y = "Number of Upsets"
  ) +
  theme_minimal() +
  geom_hline(yintercept = mean(upset_summary$num_upsets),
             linetype = "dashed", color = "red", alpha = 0.5)
```

### Explain
From the plot, we can see the number of upsets each year on the y-axis, with years ranging from 2008 to 2025 on the x-axis. The year with the fewest upsets is 2025, while 2014 has the most. The dashed red line represented the average number of upsets across all years in the dataset. In addition to the number of upsets, the color of each point represents the average seed difference in those games. For example, if a 12-seed beats a 5-seed, the seed difference is 7. Based on this, we see larger average seed differences in years like 2021 and 2013, and smaller differences in years like 2025 and 2009. 

### Describe
Looking more closely at 2025, there are not only fewer upsets, but the upsets that do occur tend to have smaller seed differences. Tournaments like this are often described as "chalky," meaning that the higher seeded teams generally perform as expected. On the other end, years like 2021 and 2022 show both a higher number of upsets and larger seed differences. One possible explanation for this is the impact of COVID-19 on recruiting and player development During this time, recruiting was more limited and disrupted, which may have led to better players ending up at smaller program, so then those smaller teams could be more likely to become Cinderella teams.

### Connect
Overall, this plot suggest that the number of upsets varies from year to year without a clear trend, which indicated that upset frequency alone may not be enough to explain unexpected success. This reinforces the idea to look deeper into team characteristics.


## Plot 2: 
- On-court performance
- Plot idea: geom_boxplot
- Data: dev --> on_court_performance --> data --> boxplot_data
  - maybe use plotly to see team name for outliers
  - also could change the metrics shown, decide the best 6 out of the 10
  - change cinderella to yes and no instead of 1 and 0
```{r}
ggplot(boxplot_data, aes(x = cinderella, y = value, fill = cinderella)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free") +
  labs(
    title = "On-Court Metrics: Cinderella vs Non-Cinderella Teams",
    x = "Cinderella",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

## Plot 3: 
- Other factors
- Plot idea: geom_boxplot
- Data: dev --> off_court_factors --> plot3_data
- maybe add a third metric into the plot, and maybe remove conference type
```{r}
ggplot(plot3_data, aes(x = cinderella, y = value, fill = conference_type)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~metric, scales = "free") +
  labs(
    title = "Off-Court Factors Behind Cinderella Runs",
    x = "Cinderella Team",
    y = "Value",
    fill = "Conference Type"
  ) +
  theme_minimal()
```

## Plot 4: 
- St. Peter's(not very good), Loyola Chicago(decent team), Davidson (good 10 seed)
- Plot idea: faceted geom_col
- Data: dev --> ... --> combined_data --> scaled_data --> plot4_data
```{r}
ggplot(plot4_data, aes(x = metric, y = value, fill = metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~team) +
  coord_flip() +
  labs(
    title = "Average Cinderella Team vs. Case Studies",
    x = "",
    y = "Standardized Value"
  ) +
  theme_minimal()
```

## Plot 5: 
- I want this plot to finish the story, maybe a regression plot or something else.
- Plot idea:
- Data: upset count, preseason ranks vs real ranks, 
```{r}
# load data & create plot

```



# Conclusion
- Tie the analysis of each plot together to determine if the amount of upsets have changed over time and why.











