---
title: "March Madness Upsets"
author: "Cooper Olney"
date: "2026-04-13"
format: 
  html:
    embed-resources: true
---

# Abstract



# Introduction
- Explain March Madness

- Define upsets & Cinderella teams, why they matter



**Main Research Question**
How have March Madness upsets evolved over time, and what factors explain the rise or decline of Cinderella runs in the modern college basketball era?

- Sub-Questions/Roadmap:
1. How has the amount of upsets in March Madness changed over time? 
2. What statistical characteristics define successful Cinderella team?
3. Case Studies: St. Peter's, Loyola Chicago, George Mason
4. Has the rise of the transfer portal and NIL changed the likelihood of upsets?


The data comes a group of datasets from Kaggle 


# Visualizations
```{r}
library(tidyverse)
```

```{r}
# Cleaning datasets

```

## Plot 1: 
- How has the amount of upsets in March Madness changed over time? 
- Plot idea: 
- Data: upset count, upset seed info, seed results, tournament matchups
```{r}
# clean data & create plot
# use upset_count dataset, then change Year into a categorical var, and look at how total number of upsets changes through the years
# could look at total or difference, maybe make two plots

```


## Plot 2: 
- What statistical characteristics define successful Cinderella team?
- Plot idea: 
- Data: shooting split, team rankings, ken bart, evan miya, 538
```{r}
# load data & create plot
# look at 3pt rate, off reb % ...
# then look at how teams are built 

```

## Plot 3: 
- St. Peter's, Loyola Chicago, George Mason, maybe other teams
- Plot idea:
- Data: team results, tournament matchups, resumes, coach results
```{r}
# load data & create plot

```

## Plot 4: 
- Has the rise of the transfer portal and NIL changed the likelihood of upsets?
- Plot idea:
- Data: upset count, preseason ranks vs real ranks, 
```{r}
# load data & create plot

```



# Conclusion
- Tie the analysis of each plot together to determine if the amount of upsets have changed over time and why.











