---
title: "March Madness Upsets"
author: "Cooper Olney"
date: "2026-04-13"
format: 
  html:
    embed-resources: true
---

# Abstract



# Introduction
- Explain March Madness

- Define upsets & Cinderella teams, why they matter

**Main Research Question**
How have March Madness upsets evolved over time, and what factors explain the rise or decline of Cinderella runs in the modern college basketball era?

- Sub-Questions/Roadmap:
1. How has the amount of upsets in March Madness changed over time? 
2. What statistical characteristics define successful Cinderella team?
3. Case Studies: St. Peter's, Loyola Chicago, George Mason
4. Has the rise of the transfer portal and NIL changed the likelihood of upsets?



# Visualizations

## Plot 1: The Rise/Fall of Upsets
- How has the amount of upsets in March Madness changed over time? 
- Plot idea: 
```{r}
# load data & create plot

```


## Plot 2: What does Cinderlla look like
- What statistical characteristics define successful Cinderella team?
- Plot idea:
```{r}
# load data & create plot

```

## Plot 3: Case Studies
- St. Peter's, Loyola Chicago, George Mason, maybe other teams
- Plot idea:
```{r}
# load data & create plot

```

## Plot 4: The Modern Era Shift
- Has the rise of the transfer portal and NIL changed the likelihood of upsets?
- Plot idea:
```{r}
# load data & create plot

```



# Conclusion
- Tie the analysis of each plot together to determine if the amount of upsets have changed over time and why.











