---
title: "March Madness Upsets"
author: "Cooper Olney"
date: "2026-04-13"
format: 
  html:
    embed-resources: true
---

# Abstract



# Introduction
The NCAA Men's Basketball Tournament, commonly known as March Madness, is one of the most widely followed and unpredictable sporting events in the United States. It features 68 teams in a single elimination format, it includes many of the most well known teams in the country and some lesser known teams. This structures creates unexpected outcomes, making March Madness especially exciting for fans filling out brackets, however no one has ever achieved a perfect bracket in the history of the tournament.

One of the most defining features of March Madness is the occurrence of upsets, in which a lower seeded teams defeats a higher seeded team, for this data the seeds need to differ by at least three spots. Higher sees are generally expected to perform better, but the tournament consistently produces surprising results. The lower seeded teams that make deep runs in the tournament are often referred to as "Cinderella teams". These teams are a major reason why March Madness remains so exciting and difficult to predict.

The data used in this analysis comes two datasets (dev & upset) sourced from Kaggle. Upset tracks the data on the number of upsets per tournament, the round the upset occurred and the seeds of the teams in those games. dev is a collection of different on the court metrics and other off the court statistics. Together, these variables off insight... (clean this up)

**Main Research Question**
What team characteristics are associated with unexpected success in March Madness?

To address this question, this analysis explores several related areas. First, it examines how common upsets are in order to provide context for how frequently Cinderella teams actually are. It then investigates which on-court performance metrics are most associated with unexpected success. In addition, the analysis considers some off-court factors like team experience and physical attributes. Finally, famous Cinderella teams like St. Peters, Loyola Chicago, and Davidson are analyzed to determine whether they follow similar patterns.

Some additional context about the structure of the tournament is important for interpreting these results. Since March Madness is single elimination it increasing the role of variability and short term performance. Additionally, games are played at neutral sites, which reduce advantages what occur in the regular season. These factors contribute to the unpredictability of outcomes and make it more challenging to identify consistent predictors of success.

However, this analysis will determine if there are patterns in the unexpected success in the tournament. Overall, these factors motivate a deeper exploration into the statistical and contextual forces of upsets and Cinderella runs in March Madness. (clean this up)


# Visualizations
```{r}
library(tidyverse)
library(readxl)
```

```{r}
# Loading kaggle datasets
upset_seed_info <- read_csv(here::here("data/mm_data_kaggle/Upset Seed Info.csv"))

dev <- read_csv(here::here("data/new_mm_datasets/DEV _ March Madness.csv"))
```

## Cleaning datasets

### Plot 1
```{r}
upset_summary <- upset_seed_info |> group_by(YEAR) |>
  summarise(num_upsets = n(),
            avg_seed_diff = round(mean(`SEED DIFF`, na.rm=TRUE), 3))
```

### Plot 2
```{r}
on_court_performance <- dev |> 
  filter(`Post-Season Tournament Sorting Index` == 1) |>
  select(
    Season, `Mapped ESPN Team Name`, Seed, 
    `Net Rating`, `Adjusted Tempo`,
    `Adjusted Offensive Efficiency`, 
    eFGPct, TOPct, ORPct, FTRate, Off3PtFG, FTPct,
    `Adjusted Defensive Efficiency`, DefFT
  ) |>
  mutate(Seed = as.numeric(Seed)) |>
  filter(Seed >= 10)

# write_csv(on_court_performance, "on_court_performance.csv")

# read in manually cleaned data from excel
data <- read_excel("data/new_mm_datasets/on_court_performance.xlsx")

data <- data |> select(-c(`Avg Possession Length (Offense)`, `Avg Possession Length (Defense)`, Bench)) |>
  relocate(cinderella, .after = Seed) |>
  mutate(cinderella = if_else(cinderella == 1, "Yes", "No"))

# plot dataset
boxplot_data <- data |>
  select(
    cinderella,
    `Net Rating`,
    `Adjusted Offensive Efficiency`,
    `Adjusted Defensive Efficiency`,
    ORPct,
    TOPct,
    `Adjusted Tempo`
  ) |>
  pivot_longer(-cinderella, names_to = "metric", values_to = "value")
```

### Plot 3
```{r}
off_court_factors <- dev |> filter(`Post-Season Tournament Sorting Index` == 1) |>
  select(
    Season, `Mapped ESPN Team Name`, Seed, `Short Conference Name`,
    AvgHeight, CenterHeight, PFHeight, SFHeight, SGHeight, PGHeight, EffectiveHeight, Experience, `Active Coaching Length` 
         ) |>
  mutate(Seed = as.numeric(Seed)) |>
  filter(Seed >= 10) |>
  filter(Season >= 2007)

# helper datasets
cinderella_var_data <- data |>
  select(Season, `Mapped ESPN Team Name`, cinderella)

off_court_factors <- off_court_factors |>
  left_join(cinderella_var_data, 
            by = c("Season", "Mapped ESPN Team Name"))

power_confs <- c("ACC", "B10", "B12", "SEC", "P12", "BE")

off_court_factors <- off_court_factors |>
  mutate(
    `Active Coaching Length` = na_if(`Active Coaching Length`, "Unknown"),
    `Active Coaching Length` = gsub(" years?| year", "", `Active Coaching Length`),
    `Active Coaching Length` = as.numeric(`Active Coaching Length`),
    conference_type = if_else(
      `Short Conference Name` %in% power_confs, "Power", "Mid-Major"
    )
  )

# plot data
plot3_data <- off_court_factors |>
  select(cinderella, conference_type, Experience, EffectiveHeight) |>
  pivot_longer(
    cols = -c(cinderella, conference_type),
    names_to = "metric",
    values_to = "value"
  )
```

### Plot 4
```{r}
combined_data <- inner_join(off_court_factors, data, 
                            by = c("Season", "Mapped ESPN Team Name", "Seed", "cinderella"))

scaled_data <- combined_data |>
  mutate(
    NetRating_z = scale(`Net Rating`),
    OffEff_z = scale(`Adjusted Offensive Efficiency`),
    DefEff_z = scale(`Adjusted Defensive Efficiency`),
    Experience_z = scale(Experience),
    EffHeight_z = scale(EffectiveHeight)
  )

avg_cinderella <- scaled_data |>
  filter(cinderella == "Yes") |>
  summarise(
    team = "Avg Cinderella",
    NetRating_z = mean(NetRating_z, na.rm = TRUE),
    OffEff_z = mean(OffEff_z, na.rm = TRUE),
    DefEff_z = mean(DefEff_z, na.rm = TRUE),
    Experience_z = mean(Experience_z, na.rm = TRUE),
    EffHeight_z = mean(EffHeight_z, na.rm = TRUE)
  )

case_teams <- scaled_data |>
  filter(
    (`Mapped ESPN Team Name` == "Davidson" & Season == 2008) |
    (`Mapped ESPN Team Name` == "Saint Peter's" & Season == 2022) |
    (`Mapped ESPN Team Name` == "Loyola Chicago" & Season == 2018)
  ) |>
  select(team = `Mapped ESPN Team Name`, 
         NetRating_z, OffEff_z, DefEff_z, Experience_z, EffHeight_z)

# plot data
plot4_data <- bind_rows(case_teams, avg_cinderella) |>
  pivot_longer(-team, names_to = "metric", values_to = "value")
```


## Plot 1: March Madness Upsets Over Time

To understand how common upsets are in March Madness, we can look at the following lollipop plot, which shows the number of upsets in each tournament.
```{r}
ggplot(upset_summary, aes(x = YEAR, y = num_upsets)) +
  geom_segment(aes(x = YEAR, xend = YEAR, y = 0, yend = num_upsets,
                   color = avg_seed_diff)) +
  geom_point(aes(color = avg_seed_diff), size = 2) +
  scale_color_viridis_c(name ="Avg Seed Diff") +
  coord_flip() +
  labs(
    title = "March Madness Upsets Over Time",
    x = "Year",
    y = "Number of Upsets"
  ) +
  theme_minimal() +
  geom_hline(yintercept = mean(upset_summary$num_upsets),
             linetype = "dashed", color = "red", alpha = 0.5)
```

### Describe
From the plot, we can see the number of upsets each year on the y-axis, with years ranging from 2008 to 2025 on the x-axis. The year with the fewest upsets is 2025, while 2014 has the most. The dashed red line represented the average number of upsets across all years in the dataset. In addition to the number of upsets, the color of each point represents the average seed difference in those games. For example, if a 12-seed beats a 5-seed, the seed difference is 7. Based on this, we see larger average seed differences in years like 2021 and 2013, and smaller differences in years like 2025 and 2009. 

### Explain
Looking more closely at 2025, there are not only fewer upsets, but the upsets that do occur tend to have smaller seed differences. Tournaments like this are often described as "chalky," meaning that the higher seeded teams generally perform as expected. On the other end, years like 2021 and 2022 show both a higher number of upsets and larger seed differences. One possible explanation for this is the impact of COVID-19 on recruiting and player development During this time, recruiting was more limited and disrupted, which may have led to better players ending up at smaller program, so then those smaller teams could be more likely to become Cinderella teams.

### Connect
Overall, this plot suggest that the number of upsets varies from year to year without a clear trend, which indicated that upset frequency alone may not be enough to explain unexpected success. This reinforces the idea to look deeper into team characteristics.


## Plot 2: On-court Metrics

To understand what on-court statistics are associated with Cinderella teams in March Madness, we can examine the following boxplot. This plot compares double-digit seeded teams that became Cinderella teams to those that did not.
```{r}
ggplot(boxplot_data, aes(x = cinderella, y = value, fill = cinderella)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free") +
  labs(
    title = "On-Court Metrics: Cinderella vs Non-Cinderella Teams",
    x = "Cinderella",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

### Describe
The plot compares double-digit seeded teams across six difference metrics that are commonly used to evaluate basketball performance. These include metrics such as Adjusted Defensive Efficiency, Adjusted Offensive Efficiency, and Net Rating. Looking at the distributions, Cinderella teams tend to have higher Adjusted Offensive Efficiency and a noticeably higher Net Rating compared to non-Cinderella teams. However, somewhat surprisingly, they appear to have a worse Adjusted Defensive Efficiency. For the remaining metrics, the differences between the two groups are minuscule.

### Explain
One possible explanation for the higher Net Rating is that Cinderella teams are more likely to come from the stronger end of the double-digit seed range, such as 10, 11, or 12 seeds, rather than 15 or 16 seeds. Since these teams are generally stronger overall, this could help explain their higher Net Rating. More importantly, the results suggest that offensive performance may play a larger role than defense in these unexpected tournament runs. teams that are highly efficient offensively can "get hot" and score at a high level, which is especially impactful in a single elimination setting where one strong performance can be enough to advance. Even if these teams are weaker defensively, a strong offense may allow them to win multiple games. 

### Connect
Overall, this plot suggests that for double-digit seeded teams, offensive strength is more strongly associated with making a Cinderella run than defensive performance. This provides initial evidence that certain on-court characteristics, particularly those related to scoring efficiency may help explain unexpected success in March Madness. To build on this, a deeper look into more specific offensive metrics could provide further insight into what drives this unexpected success.


## Plot 3:

To understand what off-court factors may influence unexpected success in March Madness, we can examine the following boxplot. This plot compares double-digit seeded teams that became Cinderella teams to those that did not, across several off-court indicators.
```{r}
ggplot(plot3_data, aes(x = cinderella, y = value)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~metric, scales = "free") +
  labs(
    title = "Off-Court Factors Behind Cinderella Runs",
    x = "Cinderella Team",
    y = "Value",
    fill = "Conference Type"
  ) +
  theme_minimal()

plot3_data |> group_by(conference_type) |> summarise(n())
```

### Describe
The plot compares teams across a few different factors, including effective height and team experience. Effective height represents the average height of players on the roster weighted by minutes played, where a value of 0 corresponds to the national average and positive values indicate above average size. Team experience measures the average number of years the starting five players have been in college, where a value of 0 represents a starting lineup made entirely of freshmen.

Looking at the distributions, there are relatively small differences between Cinderella and non-Cinderella teams across these two factors. For effective height, the median slightly increases for teams that make Cinderella runs. For experience, there is no clear advantage for more experienced teams, which might be expected. In fact, Cinderella teams appear to be slightly less experienced on average.

### Explain
One possible explanation for the height result is that size can provide advantages in areas like rebounding and defense, which may help double-digit seeded teams compete with stronger opponents. However, the large overlap in the IQR suggests that effective height alone is not a reliable predictor of unexpected success. The results for experience are more surprising, as it is often assumed that older, more experienced teams perform better in high pressure situations like March Madness. Instead, these results suggest that younger trams may be just as capable of making unexpected runs.

### Connect
Overall, this plot suggest that these off-court factors do not have as strong of a relationship with Cinderella runs as the on-court metrics explored in the previous analysis. While there are small differences, they are not consistent enough to clearly explain unexpected success. This reinforces the idea that performance based metrics may play a more important role. It also suggests that other unmeasured factors, such as game location, matchup dynamics, or randomness in a single elimination format could also contribute to these outcomes.


## Plot 4: 
- St. Peter's(not very good), Loyola Chicago(decent team), Davidson (good 10 seed)
- Plot idea: faceted geom_col
- Data: dev --> ... --> combined_data --> scaled_data --> plot4_data
```{r}
ggplot(plot4_data, aes(x = metric, y = value, fill = metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~team) +
  coord_flip() +
  labs(
    title = "Average Cinderella Team vs. Case Studies",
    x = "",
    y = "Standardized Value"
  ) +
  theme_minimal()
```



# Conclusion
- Tie the analysis of each plot together to determine if the amount of upsets have changed over time and why.











