---
title: "March Madness Upsets"
author: "Cooper Olney"
date: "2026-04-13"
format: 
  html:
    embed-resources: true
---

# Abstract



# Introduction
- Explain March Madness

- Define upsets & Cinderella teams, why they matter



**Main Research Question**
What team characteristics are associated with unexpected success in March Madness?

- Sub-Questions/Roadmap:
1. How common are upsets? (provides context)
2. What on-court stats matter?
3. What off-court factors matter?
4. Case Studies: Do famous teams fit the pattern? ex. St. Peter's, Loyola Chicago, George Mason
5. Can we predict it?


The data comes a group of datasets from Kaggle 


# Visualizations
```{r}
library(tidyverse)
```

```{r}
# Cleaning datasets

# for plot 1
upset_summary <- upset_seed_info |> group_by(YEAR) |>
  summarise(num_upsets = n(),
            avg_seed_diff = round(mean(`SEED DIFF`, na.rm=TRUE), 3))

# for plot 2
on_court_performance <- dev |> filter(`Post-Season Tournament Sorting Index` == 1) |>
  select(
    Season, `Mapped ESPN Team Name`, Seed, 
    `Net Rating`, `Adjusted Tempo`,
    `Adjusted Offensive Efficiency`, `Avg Possession Length (Offense)`, eFGPct, TOPct, ORPct, FTRate, Off3PtFG, FTPct, Bench, 
    `Adjusted Defensive Efficiency`, `Avg Possession Length (Defense)`, DefFT, Off3PtFG, 
         ) |>
  mutate(Seed = as.numeric(Seed)) |>
  filter(Seed >= 10) |>
  mutate(cinderella = if_else(sweet_16 == 1, 1, 0))
  # create a cinderella vs non-cinderella team so 10>= and advanced to sweet 16
  
# create cinderella var first
plot_data <- on_court_performance |>
  select(
    cinderella, `Net Rating`, `Adjusted Offensive Efficiency`, `Adjusted Defensive Efficiency`, eFGPct, ORPct, TOPct, FTRate, Off3PtFG, FTPct, `Adjusted Tempo`
  )

summary_data <- plot_data %>%
  group_by(cinderella) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

long_data <- summary_data %>%
  pivot_longer(-cinderella, names_to = "metric", values_to = "value")

```

## Plot 1: 
- How likely are upsets?
- Plot idea: lollipop plot, x=year, n=number of upsets, color=diff
- Data: upset seed info --> upset_summary
```{r}
ggplot(upset_summary, aes(x = YEAR, y = num_upsets)) +
  geom_segment(aes(x = YEAR, xend = YEAR, y = 0, yend = num_upsets,
                   color = avg_seed_diff)) +
  geom_point(aes(color = avg_seed_diff), size = 2) +
  scale_color_viridis_c(name ="Avg Seed Diff") +
  coord_flip() +
  labs(
    title = "March Madness Upsets Over Time",
    x = "Year",
    y = "Number of Upsets"
  ) +
  theme_minimal() +
  geom_hline(yintercept = mean(upset_summary$num_upsets),
             linetype = "dashed", color = "red", alpha = 0.5)
```


## Plot 2: 
- On-court performance
- Plot idea: geom_tile
- Data: dev --> on_court_performance --> other smaller datasets
```{r}
ggplot(long_data, aes(x = cinderella, y = metric, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "On-Court Profile of Cinderella vs Other Teams",
    x = "",
    y = "Metric",
    fill = "Average Value"
  ) +
  theme_minimal()

```

## Plot 3: 
- St. Peter's, Loyola Chicago, George Mason, maybe other teams
- Plot idea:
- Data: team results, tournament matchups, resumes, coach results
```{r}
# load data & create plot

```

## Plot 4: 
- Has the rise of the transfer portal and NIL changed the likelihood of upsets?
- Plot idea:
- Data: upset count, preseason ranks vs real ranks, 
```{r}
# load data & create plot

```



# Conclusion
- Tie the analysis of each plot together to determine if the amount of upsets have changed over time and why.











